---
title: "Airbnb_NYC_2019_Analysis_with_R"
author: "Mr Ndolo"
date: "2024-12-15"
output: html_document
---

```{r warning=FALSE}
```

## Introduction
In this analysis, I will explore the 2019 Airbnb listings data for New York City. My main goals are:

* To clean and structure the dataset for accurate analysis.

* To generate insights about Airbnb listings through visualizations.

* To build predictive models to understand factors influencing price and customer behaviors

I aim to identify trends in room types, pricing, and availability across neighborhoods. Additionally, I will use machine learning models to predict prices and group listings into clusters.

## **1. Data Cleaning and Restructuring**

### **1.1 Loading and Exploring the Dataset**
```{r message=FALSE}
# Load necessary libraries
library(tidyverse)
library(lubridate)

# Import the dataset
airbnb_data <- read.csv("AB_NYC_2019.csv")

# Explore the dataset
glimpse(airbnb_data)
summary(airbnb_data)
dim(airbnb_data)
head(airbnb_data)
```
The dataset contains information about Airbnb listings in New York City, such as prices, neighborhoods, room types, and availability. By checking the structure and summary, I can identify missing values, data types, and any anomalies that need attention

### **1.2 Handling N/A, null and Missing Values**
```{r}
# Check for missing values
colSums(is.na(airbnb_data))

# Impute missing values in `reviews_per_month` with median grouped by neighborhood group
airbnb_data <- airbnb_data %>%
  group_by(neighbourhood_group) %>%
  mutate(reviews_per_month = ifelse(is.na(reviews_per_month), 
                                    median(reviews_per_month, na.rm = TRUE), 
                                    reviews_per_month)) %>%
  ungroup()

# Drop rows with missing `name` or `host_name`
airbnb_data <- airbnb_data %>% drop_na(name, host_name)

# Verify missing values resolved
colSums(is.na(airbnb_data))
```
I identified missing values in columns like reviews_per_month. Since this field is crucial for analyzing customer engagement, I imputed missing values with the median grouped by neighbourhood_group. 

Rows with missing name or host_name were removed to ensure meaningful analysis.

I feel it is now safe to continue with my analysis since I have dealt properly with the N/A values.

### **1.3 Removing Duplicates**
```{r}
# Check and remove duplicates
sum(duplicated(airbnb_data))
airbnb_data <- airbnb_data %>% distinct()
```

Duplicate entries can skew my analysis. By removing them, I ensure that each listing is unique.

Above, I can confirm that I have removed/there are no duplicate values in my analysis so my dataset will be valid for analysis.

### **1.4 Handling Outliers**
```{r}
# Identify outliers in price using IQR
Q1 <- quantile(airbnb_data$price, 0.25)
Q3 <- quantile(airbnb_data$price, 0.75)
IQR <- Q3 - Q1

# Filter to remove outliers
airbnb_data <- airbnb_data %>% 
  filter(price >= (Q1 - 1.5 * IQR) & price <= (Q3 + 1.5 * IQR))

# Boxplot to visualize outliers
ggplot(airbnb_data, aes(y = price)) + 
  geom_boxplot() +
  labs(title = "Price Distribution After Outlier Removal")
```

Prices that deviate significantly from the normal range (outliers) could mislead the analysis. 

Using the interquartile range (IQR) method, I removed the extreme values. The boxplot above confirms the filtered dataset.

I have now successfully handled the outlier and I can proceed with my analysis.

### **1.5 Restructuring the Data**
```{r}
# Convert last_review to Date type
airbnb_data$last_review <- as.Date(airbnb_data$last_review)
glimpse(airbnb_data)
```

To work well with time-based analysis, I converted the last_review column to a date format.

### **1.6 Final structure of the cleaned data**
```{r}
glimpse(airbnb_data)
summary(airbnb_data)
dim(airbnb_data)
head(airbnb_data)
```

This now proves that my data is clean and I can continue to perform my analysis on the data. I have successfully eliminated the nulls, duplicates and outliers from my data.


## **2. Data Visualization**

### **2.1 Insights Using ggplot2**
```{r}
# Distribution of room types
ggplot(airbnb_data, aes(x = room_type)) +
  geom_bar(fill = "skyblue") +
  labs(title = "Room Type Distribution", x = "Room Type", y = "Count")

# Percentage of each room type
room_type_distribution <- airbnb_data %>%
  group_by(room_type) %>%
  summarise(count = n()) %>%
  mutate(percentage = round(count / sum(count) * 100, 1))
ggplot(room_type_distribution, aes(x = "", y = percentage, fill = room_type)) +
  geom_bar(stat = "identity", width = 1) +
  coord_polar("y", start = 0) +
  geom_text(aes(label = paste0(percentage, "%")), 
            position = position_stack(vjust = 0.5), 
            color = "white", size = 4) +
  labs(title = "Distribution of Listings by Room Type", fill = "Room Type") +
  theme_void() +
  theme(plot.title = element_text(hjust = 0.5))


# Average price by neighbourhood group
ggplot(airbnb_data, aes(x = neighbourhood_group, y = price, fill = neighbourhood_group)) +
  geom_boxplot() +
  labs(title = "Price by Neighbourhood Group", x = "Neighbourhood Group", y = "Price")

# Availability of listings across neighbourhoods
ggplot(airbnb_data, aes(x = availability_365)) +
  geom_histogram(binwidth = 30, fill = "green") +
  labs(title = "Availability of Listings", x = "Availability (Days)", y = "Count")
```


### Summary of the intepretation of the above graphs

1. **Room Type Distribution:** Most listings fall under specific room types, indicating customer preferences.

2. **Price by Neighborhood Group:** Price variations reveal the most expensive areas for Airbnb listings.

3. **Availability of Listings:** Shows the range of days listings are available annually.

### **2.2 Advanced Visualizations**
```{r}
# Relationship between price and reviews
ggplot(airbnb_data, aes(x = number_of_reviews, y = price, color = neighbourhood_group)) +
  geom_point(alpha = 0.5) +
  labs(title = "Price vs. Number of Reviews", x = "Number of Reviews", y = "Price")

# Popular room types by neighbourhood group
ggplot(airbnb_data, aes(x = neighbourhood_group, fill = room_type)) +
  geom_bar(position = "fill") +
  labs(title = "Room Types by Neighbourhood Group", x = "Neighbourhood Group", y = "Proportion")
```

In the Price vs. Number of Reviews plot, the scatter plot visually explores the relationship between the price of Airbnb listings and the number of reviews they have received, with points colored according to their neighborhood group. We can see from the plot that the Queens area has the highes review where the prices were relatively low, which concludes that the larger population of AirBnB users generally preferred that place. 

In the Room Types by Neighbourhood Group plot, we can conclude that the overall least popular room type was a shared room. This is definitely going to be a bad decision for someone looking to venture into the AirBnB business.

The private rooms are popular within the cities of Bronx, Brooklyn, Queens and Staten Island.

For Manhattan, the Entire home/apartment is the most popular choice.

## **3. Machine Learning Prediction**

### **3.1 Regression Analysis**
```{r}
# Predicting price using linear regression
model <- lm(price ~ room_type + neighbourhood_group + reviews_per_month + availability_365, data = airbnb_data)
summary(model)
```

In this regression analysis, I tried to predict how factors like room type and availability influence price. The model helps identify key contributors to price changes.

### **3.2 Clustering**
```{r}
# Clustering neighborhoods based on reviews and price
cluster_data <- airbnb_data %>% select(number_of_reviews, price) %>% na.omit()
kmeans_result <- kmeans(cluster_data, centers = 3)
airbnb_data$cluster <- as.factor(kmeans_result$cluster)

# Visualize clusters
ggplot(cluster_data, aes(x = number_of_reviews, y = price, color = as.factor(kmeans_result$cluster))) +
  geom_point(alpha = 0.5) +
  labs(title = "Clustering of Listings", x = "Number of Reviews", y = "Price")
```

In this part, I have performed clustering on neighborhoods based on the number of reviews and price of Airbnb listings using the K-means clustering algorithm and then visualized the resulting clusters.

I have clustered the Airbnb listings into three groups based on the number of reviews and price, and I have assigned each listing to a cluster, then visualized these clusters in a scatter plot. 

The plot gives a good visual inspection of how the clusters relate to the variables of interest (number of reviews and price)

### **3.3 Model Evaluation**
```{r}
# Evaluate regression model using RMSE
library(Metrics)
predicted <- predict(model, airbnb_data)
actual <- airbnb_data$price
rmse(actual, predicted)
```

Here, I am evaluating the performance of a regression model by calculating the Root Mean Square Error (RMSE) between the predicted values and the actual values of the dataset.

I have simply evaluated the regression model by comparing its predictions on the Airbnb dataset against the actual prices, therefore providing a quantitative measure (RMSE) of the model's accuracy.

## **Logistic Regression Analysis**

### **1. Problem Definition**
In this analysis, I aim to predict whether an Airbnb listing is **affordable** (binary outcome variable: 1 = Affordable, 0 = Expensive) based on the following predictor variables:
- **Room Type**: Type of room offered (categorical variable).
- **Minimum Nights**: Minimum nights required for booking (numerical variable).

**Objective**: I will Evaluate how these factors influence affordability and use logistic regression to predict affordability. I hypothesize that both predictors are statistically significant in determining affordability.

### **2. Data Preparation**
```{r, echo=TRUE, message=FALSE, warning=FALSE}
# Load necessary libraries
library(tidyverse)
library(caret)

# Load dataset
airbnb_data <- read.csv("AB_NYC_2019.csv")

# Create the binary outcome variable (affordable: 1 if price < median, 0 otherwise)
median_price <- median(airbnb_data$price, na.rm = TRUE)
airbnb_data <- airbnb_data %>%
  mutate(affordable = ifelse(price < median_price, 1, 0))

# Select relevant variables
data_model <- airbnb_data %>%
  select(affordable, room_type, minimum_nights) %>%
  filter(!is.na(affordable))

# Check for correlations
correlation_matrix <- cor(data_model %>% select(minimum_nights), use = "complete.obs")
correlation_matrix
```

To begin, I prepared the data by creating a binary variable affordable, which categorizes listings as either affordable (1) or expensive (0) based on whether their price is below the median price. We also selected room_type and minimum_nights as the predictor variables for the analysis. These variables were chosen because room type is likely to influence pricing, and minimum nights could reflect booking constraints that affect affordability.

I also checked for correlations among the predictors to ensure they are not highly correlated, as multicollinearity can negatively impact the logistic regression model.

### **3. Exploratory Data Analysis**
```{r, echo=TRUE, message=FALSE, warning=FALSE}
# Distribution of affordability
ggplot(data_model, aes(x = factor(affordable), fill = factor(affordable))) +
  geom_bar() +
  labs(title = "Distribution of Affordable Listings", 
       x = "Affordable (1 = Yes, 0 = No)", 
       y = "Count") +
  scale_fill_manual(values = c("0" = "red", "1" = "green"), name = "Affordable") +
  theme_minimal()

# Minimum nights by affordability
ggplot(data_model, aes(x = affordable, y = minimum_nights)) +
  geom_boxplot(aes(fill = factor(affordable))) +
  scale_fill_manual(values = c("0" = "red", "1" = "green"), name = "Affordable") +
  labs(title = "Minimum Nights by Affordability", x = "Affordable", y = "Minimum Nights") +
  theme_minimal()
```

Here, I performed an exploratory data analysis to understand the distribution of the outcome variable and the relationship between predictors and affordability. Visualizing these distributions helps validate our choice of predictors and offers insights into potential trends.

- **Distribution of Affordable Listings**: This bar plot shows the number of listings categorized as affordable versus expensive.
- **Minimum Nights by Affordability**: The box plot provides a comparison of the minimum nights required for affordable and expensive listings.