---
title: "Airbnb_NYC_2019_Analysis_with_R"
author: "Mr Ndolo"
date: "2024-12-15"
output: html_document
---

```{r warning=FALSE}
```

## Introduction
In this analysis, I will explore the 2019 Airbnb listings data for New York City. My main goals are:

* To clean and structure the dataset for accurate analysis.

* To generate insights about Airbnb listings through visualizations.

* To build predictive models to understand factors influencing price and customer behaviors

I aim to identify trends in room types, pricing, and availability across neighborhoods. Additionally, I will use machine learning models to predict prices and group listings into clusters.

## **1. Data Cleaning and Restructuring**

### **1.1 Loading and Exploring the Dataset**
```{r message=FALSE}
# Load necessary libraries
library(tidyverse)
library(lubridate)

# Import the dataset
airbnb_data <- read.csv("AB_NYC_2019.csv")

# Explore the dataset
glimpse(airbnb_data)
summary(airbnb_data)
dim(airbnb_data)
head(airbnb_data)
```
The dataset contains information about Airbnb listings in New York City, such as prices, neighborhoods, room types, and availability. By checking the structure and summary, I can identify missing values, data types, and any anomalies that need attention

### **1.2 Handling N/A, null and Missing Values**
```{r}
# Check for missing values
colSums(is.na(airbnb_data))

# Impute missing values in `reviews_per_month` with median grouped by neighborhood group
airbnb_data <- airbnb_data %>%
  group_by(neighbourhood_group) %>%
  mutate(reviews_per_month = ifelse(is.na(reviews_per_month), 
                                    median(reviews_per_month, na.rm = TRUE), 
                                    reviews_per_month)) %>%
  ungroup()

# Drop rows with missing `name` or `host_name`
airbnb_data <- airbnb_data %>% drop_na(name, host_name)

# Verify missing values resolved
colSums(is.na(airbnb_data))
```
I identified missing values in columns like reviews_per_month. Since this field is crucial for analyzing customer engagement, I imputed missing values with the median grouped by neighbourhood_group. 

Rows with missing name or host_name were removed to ensure meaningful analysis.

I feel it is now safe to continue with my analysis since I have dealt properly with the N/A values.

### **1.3 Removing Duplicates**
```{r}
# Check and remove duplicates
sum(duplicated(airbnb_data))
airbnb_data <- airbnb_data %>% distinct()
```

Duplicate entries can skew my analysis. By removing them, I ensure that each listing is unique.

Above, I can confirm that I have removed/there are no duplicate values in my analysis so my dataset will be valid for analysis.



